### 2023-07-27 python_excel_ite.tex
code chunks converted to screenshots

We start with a very simple model that ignores the uncertainty in demand and that is not object oriented. Proceeding much like we would in Excel, base input values are stored in variables.

\begin{tcolorbox}
\begin{minted}[gobble=0]{python}
# Base inputs
unit_cost = 7.50
selling_price = 10.00
unit_refund = 2.50

# Demand parameters
demand_mean = 193
demand_sd = 40

# Deterministic model
demand = demand_mean

# Initial value for order quantity - this is the decision variable
order_quantity = 200
\end{minted}
\end{tcolorbox}

Students are given skeleton code as scaffolding and must finish the expressions for computing key intermediate outputs (\code{order\_cost, sales\_revenue, refund\_revenue}) and the final output, \code{profit}.

\begin{tcolorbox}
\begin{minted}[gobble=0]{python}
order_cost = unit_cost * order_quantity
sales_revenue = ??? * selling_price
refund_revenue = ??? * unit_refund
profit = sales_revenue + refund_revenue - order_cost
\end{minted}
\end{tcolorbox}

@@

At this point we discuss a fundamental difference between the computing model of Excel and that of Jupyter notebooks -- while spreadsheets respond automatically to changes, Jupyter notebooks require rerunning all code cells that include or follow the changed code. This difference in behavior between Jupyter notebooks and spreadsheets might be one of the most important concepts for students to internalize. A well known challenge, often referred to as the \textit{hidden state problem}, associated with computational notebooks is the ease of running code cells out of order and introducing inadvertent errors \cite{\cite{johnson2020benefits},grusDonNotebooks2018}. This point along with strategies to avoid it are discussed during the first week of the course which focuses on \href{http://www.sba.oakland.edu/faculty/isken/courses/mis6900/jupyter_notebooks.html}{notebook computing best practices as well as a historical look} at computational notebooks.

@@

A natural next step is to do the equivalent of an Excel 2-way Data Table. By creating a profit function that takes all of the base inputs as arguments, a list comprehension can then be used to generate the equivalent of a 2-way Data Table. Even better, this actually allows us to do the equivalent of an n-way Data Table with an arbitrary number of outputs -- something Excel does not have. This example provides an effective way to show the power of creating reusable functions and Python's handy list comprehension construct.

This code chunk,

\begin{tcolorbox}[left=0mm]
\begin{minted}[gobble=1]{python}
	def bookstore_profit(unit_cost, selling_price, 
	                     unit_refund, order_quantity, demand):
	'''
	Compute profit in bookstore model
	'''
		order_cost = unit_cost * order_quantity
		sales_revenue = np.minimum(order_quantity, demand) * selling_price
		refund_revenue = np.maximum(0, order_quantity - demand)
		profit = sales_revenue + refund_revenue - order_cost
		return profit

	# Set up input vector ranges	
	demand_range = np.arange(50, 301, 5)
	order_quantity_range = np.arange(50, 301, 25)

	# Create data table (as a list of tuples)
	data_table_1 = [(d, oq, bookstore_profit(
	                 unit_cost, selling_price, unit_refund, oq, d)) 
	for d in demand_range for oq in order_quantity_range]

	# Convert to dataframe
	dtbl_1_df = pd.DataFrame(data_table_1, 
	                         columns=['Demand', 'OrderQuantity', 'Profit'])

\end{minted}
\end{tcolorbox}

results in the following output:

\begin{tcolorbox}
	\begin{verbatim}
	   Demand  OrderQuantity  Profit
	      50             50   125.0
	      50             75   -37.5
	      50            100  -200.0
	      50            125  -362.5
	      50            150  -525.0
	      ...
	     300            275   687.5
	     300            300   750.0
\end{verbatim}
\end{tcolorbox}

While laid out differently, this is the same output produced by a 2-way Data Table in Excel. At this point, we pause to consider a higher level design question.

@@

It is not clear whether this non-object oriented approach will end up making the most sense for doing spreadsheet type modeling in Python. This uncertainty is made explicit in the notebook and students are led on a bit of a journey of discovery as we explore different software designs for this modeling problem. I purposefully designed the notebooks in this way to show students that the path to a solution is not necessarily linear and that giving yourself the freedom to explore alternatives, many of which will be dead ends, can lead to a rich learning experience. It is also more realistic in the sense that this is how most real problems are solved. It is analogous to the idea that mathematical proofs usually do not reflect the myriad of failed approaches that were tried but that ultimately contributed to the final product.

@@

For our initial design of a \code{BookstoreModel} class we make all of the base inputs class attributes (properties) and then add method attributes to compute outputs such as costs, revenues and profits. Students can easily see the mapping between elements in the initial procedural model and this new object oriented version. Several important concepts and syntactical details of doing object oriented programming Python are highlighted. Again, the goal is to weave in these more advanced Python programming concepts within the context of a very familiar modeling problem. This section of the notebook ends with using the \code{BookstoreModel} class to create a new model object instance with all of its base inputs instantiated with values. Before moving, on students are presented with a few challenges involving enhancements to the \code{BookstoreModel} class. Interspersing short coding challenges throughout the notebook gives students a chance to test themselves on their understanding and to get some coding practice.

A few design dilemmas are posed and some non-working approaches illustrated as we try to decide on how to implement an n-way data table function for this new object oriented model. These meta reflections are intentionally included to better mimic actual software development and reinforce the idea that software development is much more than simply writing code to implement a perfectly thought out design. We forge ahead with some preliminary ideas and are confronted with an intermediate problem related to generating a list of dictionaries that represent combinations of inputs, or \textit{scenarios}, to evaluate. This provides a great opportunity to remind students of one of the great strengths of the open source software ecosystem -- we can leverage work done in other packages and even look at actual source code to see how something was implemented. In this case, the well known \href{https://scikit-learn.org/stable/}{scikit-learn} package \cite{scikit-learn} has a \code{ParameterGrid} function that solves our scenario generation problem quite nicely. This example also reminds the students of the value of being able to make sense of API documentation and of object oriented code written by others.

@@

\begin{figure}[!ht]
	\centering
	\begin{minipage}{.80\textwidth}
		\includegraphics[width=1.0\textwidth]{images/two_way_dt.png}
	\end{minipage}
	\caption{Plot based on 2-way data table}
	\label{fig:2way_dt}
\end{figure}

@@

In the Excel modeling world, Monte-Carlo simulation can be done without add-ins, but packages like @Risk \cite{palisadeRISKRiskAnalysis} can make the process much easier. In particular, random variable generation for a wide range of probability distributions is facilitated by add-ins like @Risk. In Python, the NumPy package provide random variate generation for a large number of distributions and similarly, SciPy provides functions for computing distribution related quantities for many probability distributions. Much like we did when creating the \code{data\_table} function, we start with adding uncertainty to a single input variable in the bookstore model and rely on NumPy's vectorized computing capabilities to generate the profit associated with each realization of the random input. Then we can do standard statistical analysis of the simulation output using Python packages such as pandas and SciPy. In spreadsheet based simulation textbooks, the notion of the ``Flaw of Averages'' popularized by Sam Savage \cite{savageFlawAveragesWhy2012} is often illustrated. We do the same with our Python based model, showing how replacing random demand by its mean can result in a wildly optimistic estimate of the mean profit found through analyzing the simulation output. Much like doing simulation in Excel without add-ins, using Python makes transparent the quite simple and brute force nature of Monte-Carlo simulation.

>>>>

We start with a very simple model that ignores the uncertainty in demand and that is not object oriented. Proceeding much like we would in Excel, base input values are stored in variables. Students are given skeleton code as scaffolding and must finish the expressions for computing key intermediate outputs (\code{order\_cost, sales\_revenue, refund\_revenue}) and the final output, \code{profit} -- see Figure \ref{fig:model_inputs}

@@

At this point we discuss a fundamental difference between the computing model of Excel and that of Jupyter notebooks -- while spreadsheets respond automatically to changes, Jupyter notebooks require rerunning all code cells that include or follow the changed code. This difference in behavior between Jupyter notebooks and spreadsheets might be one of the most important concepts for students to internalize. A well known challenge, often referred to as the \textit{hidden state problem}, associated with computational notebooks is the ease of running code cells out of order and introducing inadvertent errors \cite{johnson2020benefits,grusDonNotebooks2018}. This point along with strategies to avoid it are discussed during the first week of the course which focuses on \href{http://www.sba.oakland.edu/faculty/isken/courses/mis6900/jupyter_notebooks.html}{notebook computing best practices as well as a historical look} at computational notebooks.

@@

A natural next step is to do the equivalent of an Excel 2-way Data Table. By creating a profit function that takes all of the base inputs as arguments, a list comprehension can then be used to generate the equivalent of a 2-way Data Table. Even better, this actually allows us to do the equivalent of an n-way Data Table with an arbitrary number of outputs -- something Excel does not have. This example provides an effective way to show the power of creating reusable functions and Python's handy list comprehension construct. As shown in Figure \ref{fig:function_datatable}, a few code and markdown cells are used to create the code, provide some explanation and show the output. This is the same output one would get with an Excel 2-way Data Table, just laid out differently. Then in Figure \ref{fig:datatable_scatter} we show how the output can be turned into a scatter plot with profit mapped to a color hue.

\begin{figure*}[!htbp]
	\centering
	\caption{A 2-way data table}
	\includegraphics[scale=0.5]{images/function_datatable.png}
	\label{fig:function_datatable}
\end{figure*}

\begin{figure*}[!htbp]
	\centering
	\caption{Scatter plot based on 2-way data table}
	\includegraphics[scale=0.5]{images/datatable_scatter.png}
	\label{fig:datatable_scatter}
\end{figure*}

@@

It is not clear whether this non-object oriented procedural approach will end up making the most sense for doing spreadsheet type modeling in Python. This uncertainty is made explicit in the notebook and students are led on a bit of a journey of discovery as we explore different software designs for this modeling problem. I purposefully designed the notebooks in this way to show students that the path to a solution is not necessarily linear and that giving yourself the freedom to explore alternatives, many of which will be dead ends, can lead to a rich learning experience. It is also more realistic in the sense that this is how most real problems are solved. It is analogous to the idea that mathematical proofs usually do not reflect the myriad of failed approaches that were tried but that ultimately contributed to the final product.

@@

For our initial design of a \code{BookstoreModel} class we make all of the base inputs class attributes (properties) and then add method attributes to compute outputs such as costs, revenues and profits. Students can easily see the mapping between elements in the initial procedural model and this new object oriented version. Several important concepts and syntactical details of doing object oriented programming Python are highlighted. Again, the goal is to weave in these more advanced Python programming concepts within the context of a very familiar modeling problem. This section of the notebook ends with using the \code{BookstoreModel} class to create a new model object instance with all of its base inputs instantiated with values. Before moving on, students are presented with a challenges involving enhancements to the \code{BookstoreModel} class. Interspersing short coding challenges throughout the notebook gives students a chance to test themselves on their understanding and to get some coding practice.

A few design dilemmas are posed and some non-working approaches illustrated as we try to decide on how to implement an n-way data table function for this new object oriented model. These meta reflections are intentionally included to better mimic actual software development and reinforce the idea that software development is much more than simply writing code to implement a perfectly thought out design. We forge ahead with some preliminary ideas and are confronted with an intermediate problem related to generating a list of dictionaries that represent combinations of inputs, or \textit{scenarios}, to evaluate. This provides a great opportunity to remind students of one of the great strengths of the open source software ecosystem -- we can leverage work done in other packages and even look at actual source code to see how something was implemented. In this case, the well known \href{https://scikit-learn.org/stable/}{scikit-learn} package \cite{scikit-learn} has a \code{ParameterGrid} function that solves our scenario generation problem quite nicely. This example also reminds the students of the value of being able to make sense of API documentation and to understand object oriented code written by others.

@@

\begin{figure}[!ht]
	\centering
	\begin{minipage}{.80\textwidth}
		\includegraphics[width=1.0\textwidth]{images/two_way_dt.png}
	\end{minipage}
	\caption{Faceted plot based on 2-way data table}
	\label{fig:2way_dt}
\end{figure}

@@

In the Excel modeling world, Monte-Carlo simulation can be done without add-ins, but packages like \href{https://lumivero.com/products/at-risk/}{@Risk} can make the process much easier. In particular, random variable generation for a wide range of probability distributions is facilitated by add-ins like @Risk. In Python, the \href{https://numpy.org/doc/stable/reference/random/generator.html}{NumPy package provide random variate generation} for a large number of distributions and similarly, \href{https://docs.scipy.org/doc/scipy/reference/stats.html}{SciPy provides functions for computing distribution related quantities} for many probability distributions. Much like we did when creating the \code{data\_table} function, we start with adding uncertainty to a single input variable in the bookstore model and rely on NumPy's vectorized computing capabilities to generate the profit associated with each realization of the random input. Then we can do standard statistical analysis of the simulation output using Python packages such as pandas and SciPy. In spreadsheet based simulation textbooks, the notion of the ``Flaw of Averages'' popularized by Sam Savage \cite{savageFlawAveragesWhy2012} is often illustrated. We do the same with our Python based model, showing how replacing random demand by its mean can result in a wildly optimistic estimate of the mean profit found through analyzing the simulation output. Much like doing simulation in Excel without add-ins, using Python makes transparent the quite simple and brute force nature of Monte-Carlo simulation.
